---
title: "Linear Regression with NYS Report Card"
author: "Jason M. Bryer, Ph.D."
date: "November 17, 2014"
output: 
  slidy_presentation:
    keep_md: true
---

## Data Preparation

The dataset contains grade 8 math scores for Albany County
 
```{r}
load('../Data/NYSReportCardAlbany.Rda')
rc.albany <- rc.albany[,c('School','Pass2012','Pass2013')]
print(rc.albany, row.names=FALSE)
```

## What we wish to estimate:

$$Pass_{2013} = {b}_{1}Pass_{2012} + {b}_{0}$$

```{r, scatterplot}
plot(rc.albany$Pass2012, rc.albany$Pass2013, xlim=c(0,100), ylim=c(0,100))
```

## Descriptive Statistics


$$ { mean }_{ x }=\overline { x } =\frac { \sum { { x }_{ i } }  }{ n } $$

```{r means}
(mean2012 <- mean(rc.albany$Pass2012))
(mean2013 <- mean(rc.albany$Pass2013))
```

$$ Variance_{ x }={ s }_{ x }^{ 2 }=\frac { \sum { ({ x }_{ i }-\overline { x }  } )^{ 2 } }{ n-1 }  $$

```{r variances}
var(rc.albany$Pass2012)
var(rc.albany$Pass2013)
```

$$ { Standard\quad Deviation }_{ x }=s=\sqrt { { s }^{ 2 } } $$

```{r sds}
(sd2012 <- sd(rc.albany$Pass2012))
(sd2013 <- sd(rc.albany$Pass2013))
```

Number of Rows (n)

```{r n}
(n <- nrow(rc.albany))
```

## z-Scores

$$ z=\frac { y-\overline { y }  }{ s } $$

```{r zscores}
rc.albany$Pass2012.z <- (rc.albany$Pass2012 - mean2012) / sd2012
rc.albany$Pass2013.z <- (rc.albany$Pass2013 - mean2013) / sd2013
print(rc.albany[,-1], row.names=FALSE)
```

## Correlation

$$ r=\frac { \sum { { z }_{ x }{ z }_{ y } }  }{ n-1 } $$

```{r corr}
rc.albany$ZProduct <- rc.albany$Pass2012.z * rc.albany$Pass2013.z
r <- sum( rc.albany$ZProduct ) / ( n - 1 )
r
print(rc.albany[,-1], row.names=FALSE)
```

## Correlation (cont.)

```{r corrtest}
cor.test(rc.albany$Pass2012, rc.albany$Pass2013)
```

## Slope

$$ m = r \frac{s_y}{s_x} $$

```{r slope}
m <- r * (sd2013 / sd2012)
m
```

## Intercept:

```{r intercept}
b <- mean2013 - m * mean2012
b
```

## Draw Regression Line

```{r scatterWithLine}
plot(rc.albany$Pass2012, rc.albany$Pass2013, xlim=c(0,100), ylim=c(0,100))
abline(h=mean2013, col='blue')
abline(v=mean2012, col='blue')
abline(a=b, b=m, col='red')
```

## Residuals

```{r residuals}
rc.albany$Predicted2013 <- m * rc.albany$Pass2012 + b
rc.albany$Residual <- rc.albany$Pass2013 - rc.albany$Predicted2013
print(rc.albany[,-1], row.names=FALSE)
```

## Residual Plot

```{r residualPlot}
plot(rc.albany$Pass2012, rc.albany$Residual)
abline(h=0)
```

## Histogram of Residuals

```{r residualHistogram}
hist(rc.albany$Residual)
```

## Quantile-Quantile Plot

```{r qqplot}
qqnorm(rc.albany$Residual)
qqline(rc.albany$Residual, col=2)
```

## Final Regression Line and $R^2$

```{r, echo=FALSE, results='asis'}
cat( paste0('$$ Pass_{2013} = ', round(m, digits=2), ' * Pass_{2012}', ifelse(b < 0, ' - ', ' + ') , round(abs(b), digits=2), ' $$' ))
```

```{r}
r ^ 2
```

The mean score from 2012 accounts for `r round(r^2 * 100, digits=1)`% of the variance in the mean score from 2013 in math at grade 8.


## Regression in R

```{r}
lm.out <- lm(Pass2013 ~ Pass2012, data=rc.albany)
summary(lm.out)
```