---
title: DATA 606 Data Project Proposal
author: 
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data

```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**



### Cases 

**What are the cases, and how many are there?**



### Data collection 

**Describe the method of data collection.**



### Type of study 

**What type of study is this (observational/experiment)?**



### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**




### Response 

**What is the response variable, and what type is it (numerical/categorical)?**



### Explanatory 

**What is the explanatory variable, and what type is it (numerical/categorival)?**



### Relevant summary statistics 

**Provide summary statistics relevant to your research question. For example, if you’re comparing means across groups provide means, SDs, sample sizes of each group. This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

```

